Anony*mice*d shareable data: Using *mice* to create and analyze multiply imputed synthetic data sets

### 2021-11-01 manuscript/Manuscript.Rmd

---
title: "Anony*mice*d shareable data: Using *mice* to create and analyze multiply imputed synthetic data sets"
author:
  - name: Thom Volker
    affil: 1,*, \dagger
    orcid: 0000-0002-2408-7820
  - name: Gerko Vink
    affil: 1, \dagger
    orcid: 0000-0001-9767-1924
affiliation:
  - num: 1
    address: |
      Utrecht University - 
      Department of Methodology and Statistics. 
      Padualaan 14. 3584CH Utrecht, the Netherlands
# firstnote to eighthnote
firstnote: |
  These authors contributed equally to this work.
correspondence: |
  t.b.volker@uu.nl.
journal: psych
type: article
status: submit
bibliography: synth.bib
appendix: appendix.tex
abstract: |
  Synthetic data sets can greatly improve the dissemination of data 
  and further analysis of private data. Generating and analyzing synthetic data 
  sets is straightforward, yet a synthetic data analysis pipeline is 
  seldom adopted by applied researchers. We outline a simple procedure for 
  generating and analyzing synthetic data sets with `mice` in `R`. We demonstrate 
  using simulations that the analysis results obtained on synthetic data 
  yields unbiased and valid inferences, and leads to synthetic 
  records that cannot be distinguished from the true data records. The ease of use 
  when synthesizing data with `mice`, together with the validity of inferences 
  obtained through this procedure opens up a wealth of possibilities for data 
  dissemination and further research of initially private data. 
keywords: |
  mice; imputation; synthetic data.
acknowledgement: |
  We are grateful to Mirthe Hendriks and Stijn van den Broek for replicating the 
  simulation study on an independent data set. Additionally, we thank Hanne Oberman and two anonymous reviewers for their helpful and insightful comments.
authorcontributions: |
  The authors contributed equally to this work.
conflictsofinterest: |
  The authors declare no conflict of interest. 
sampleavailability: |
  A full simulation archive is available from https://github.com/amices/Federated_imputation/tree/master/manuscript.
output: rticles::mdpi_article
---

@@

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values, one could easily replace any observed sensitive values by multiple draws from the posterior predictive distribution - or a non-Bayesian predictive distribution. We will refer to both of these as predictive distributions. Similarly to multiple imputation of missing data, the multiple synthetic data sets allow for correct statistical inferences, despite the fact that the analyses do not use the "true" value. The analyses over multiple synthetic data sets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure.

@@

The `mice` package [@mice] in `R` [@Rproject] has been developed for multiple imputation to overcome problems related to nonresponse. In that context, the aim is to replace missing values with plausible values from the predictive distribution of that variable. In `mice`, this is accomplished using fully conditional specification (FCS) [@vanbuuren_fully_2006], which breaks down the multivariate distribution of the data $\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})$ into $j = 1, 2, \dots, k$ univariate conditional densities, where $k$ denotes the number of columns in the data. Using FCS, a model is constructed for every incomplete variable and the missing values $Y_{j, mis}$ are then imputed with draws from the predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ on a variable-by-variable basis. Note that the predictor matrix $Y_{-j}$ may contain yet imputed values from an earlier imputation step, and thus will be updated after every iteration. This procedure is applied $m$ times, resulting in $m$ completed data sets $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis})$.

@@

The imputation approach in `mice` can be easily extended to generate synthetic values. Rather than imputing missing data, observed values are then replaced by synthetic draws from the predictive distribution. For simplicity, assume that the data is completely observed (i.e., $\textbf{Y} = \textbf{Y}_{obs}$). Following the notation of @reiter_raghunathan_multiple_2007, let, given $n$ observations, $Z_i = 1$ if any of the values of unit $i = 1, 2, \dots, n$, are to be replaced by imputations, and $Z_i = 0$ otherwise, with $\textbf{Z} = (Z_1, Z_2, \dots, Z_n)$. Accordingly, the data consists of values that are to be replaced and values that are to be kept (i.e., $\textbf{Y} = (\textbf{Y}_{rep}, \textbf{Y}_{nrep})$. Now, instead of imputing $\textbf{Y}_{mis}$ with draws from the predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ as in the missing data case, $\textbf{Y}_{rep}$ is imputed from the posterior distribution of $P(Y^{(l)}_{j, rep} | \textbf{Y}^{(l)}_{-j}, \textbf{Z}, \theta)$, where $l$ is an indicator for the synthetic data set ($l = 1, 2, \dots, m$). The synthetic values that are imputed at an earlier step can be used for imputing variable $j$. This process results in the synthetic data $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$. Note that, given a completely observed data set, iterating over the predictive distribution is not required.

@@

```{r eval = FALSE}
library(mice)
syn <- mice(data.to.synthesize, 
            m = 10,
            maxit = 1,
            where = matrix(TRUE,
                           nrow = nrow(data.to.synthesize),
                           ncol = ncol(data.to.synthetize)))
```

Here, the argument `where` requires a matrix of the same dimensions as the data, (i.e., a $n \times k$ matrix) containing logicals $z_{ij}$ that indicate which cells are selected to have their values replaced by draws from the predictive distribution. This approach allows to replace a subset of the observed data (e.g., by only specifying those cells that are to be replaced as `TRUE` in the `where`-matrix, leaving the rest to `FALSE`), or as in the aforementioned example, the observed data as a whole, resulting in a data set that partially or completely consists of synthetic data values. Note that because the data is completely observed, iterating over the predictive distribution is not required.

@@

Choosing an adequate imputation model to impute the data is paramount, as a flawed imputation model may drastically impact the validity of inferences [@grund2021using; @jiang2021balancing]. Imputation models should be as flexible as possible to capture most of the patterns in the data, and to model possibly unanticipated data characteristics [@murray_multiple_2018; @rubin_18years_1996]. Generally, multiple methods can be feasible, including joint modeling, sequential regression and FCS, all with different parameter specifications, depending on the imputation problem at hand [@ludtke2020regression; @yucel2018sequential; @murray2016multiple]. Various simulations have shown that the most accurate imputation procedure may vary over different imputation problems [@hanne_review_2021]. Hence, a one-size-fits-all solution does probably not exist.

However, one of the methods that has shown a reasonable and consistent performance over multiple simulations using FCS in combination with classification and regression trees [CART\; @breiman_cart_1984; @reiter_cart_2005; @burgette_reiter_cart_2010; @doove_buuren_recursive_2014; @raab_practical_2016]. Due to its non-parametric nature, CART allows to model more complex patterns (e.g., interactions and non-linear relationships) in the data. Loosely speaking, CART sequentially splits the predictor space into non-overlapping regions in such a way that the within-region variance is as small as possible after every split. As such, CART does not impose any parametric distribution on the data, making it a widely applicable method that allows for a large variety of relationships within the data [@islr_2013]. Given these appealing characteristics and the call for the use of flexible methods when multiply imputing data, we will focus our illustrations and evaluations of `mice` to method `mice.impute.cart()`, realized by:

@@

In a nutshell, the above code shows the simplicity of creating $m = 10$ synthetic data sets using `mice`. In practice, however, one should take some additional complicating factors into account. For example, one should account for deterministic relations in the data. Additionally, relations between variables may be described best using a different model than `CART`. Such factors are data dependent and should be considered by the imputer. In the next section, we will describe how the `mice::boys` data [@fredriks_boys_2000] can be adequately synthesized. Additionally, we will show through simulations that this approach yields valid inferences.

@@

Unfortunately, this data set does not differ from the vast majority of collected data sets, in the sense that it suffers from missing data. For simplicity, we complete the missing values using the default `mice` imputation model for all predictors except `bmi`, which is passively imputed using its deterministic relation with `wgt` and `hgt`. Specifically, the imputed values are used to calculate the exact `bmi` values that correspond with `hgt` and `wgt`.

@@

Usually, one would draw samples from a population that can be synthesized to evaluate the performance of the synthetisation methods. As we only have access to a single sample, 1000 bootstrap samples of the `boys` data have been synthesized with $m = 5$ imputations for every data cell to induce an appropriate amount of sampling variance. This approach precludes us from knowing the *true* data-generating model, and in this sense provides a more stringent test of the applicability of `mice` for real-life synthetisation procedures. Specifically, every bootstrapped sample is treated as an actual sample from a population that is synthesized using `mice`, using a model that is built to approximate the true data-generating mechanisms as closely as possible. As relationships in realistic data are generally more complex than relationships in parametrically simulated data, achieving good performance on a real data set is likely to be more indicative of practical applicability than good performance on a parametrically simulated data set.

@@

$$
\bar{u}_m = \sum_{l = 1} \frac{u^{(l)}}{m},
$$
where $u^{(l)}$ denotes the variance estimate in the $l$th synthetic data set. These pooling rules are implemented in `mice` as the function `pool.syn`. @reiter_kinney_2012 proved that synthesizing data from the posterior predictive distribution is not required for these pooling rules to hold. In fact, the parameters of an imputation model can be fixed to their maximum likelihood estimates or their posterior modes, and synthetic values can be drawn after plugging these values into the imputation models. This observation as been taken a step further by @raab_practical_2016, who developed pooling rules that do not require multiple synthetic data sets and performed highly similar to the aforementioned pooling rules. However, we adhere to the originally proposed pooling rules by @reiter_partially_inference_2003, who defined the total variance in terms of sampling variance and between-imputation variance.

@@

We evaluate the synthetic data with respect to the *true* data set on the basis of three aims. We believe that every reliable and valid data synthetisation effort in statistical data analysis should be able to yield 1) unbiased univariate statistical properties, 2) unbiased bivariate properties, 3) unbiased and valid multivariate inferences and 4) synthetic data that cannot be distinguished from real data. We consider the evaluation of the synthetic data simulations in the above order.

@@

@@

@@

```{r univariate-statistics-II, echo = FALSE, cache = TRUE, results = FALSE, fig.cap = "Empirical sampling distribution of the number within each category of the categorical variables in the bootstrapped and synthetic boys data."}
boot_stats_cat <- bootstrap_samples %>%
  map_dfr(~.x %>%
            select(!where(is.numeric)) %>%
            mutate(across(everything(), as.character)) %>%
            pivot_longer(everything()) %>% 
            group_by(name, value) %>%
            summarize(`Number of observations` = n(), .groups = "drop"))

@@

Both figures show that the marginal distributions of the bootstrapped data and the synthetic data are highly similar. Additionally, the univariate descriptives for the original data and the synthetic data can be found in Table 2.

```{r, include = FALSE}
library(xtable)
bind_rows(true, setNames(synth, names(true))) %>% 
           arrange(vars) %>% .[, c(2:5, 8:9, 11:12)] %>% round(3) %>% 
  xtable
```
<!-- % latex table generated in R 4.1.0 by xtable 1.8-4 package -->
<!-- % Mon Oct  4 16:47:59 2021 -->
\begin{table}[ht]
\caption{Univariate descriptives for the true data and $m=5$ pooled univariate descriptives for the synthetic data over 1000 simulations. Variable names followed by a $^*$ are categorical.}
\centering
\begin{tabular}{rrrrrrrrr}
  \hline
 & n & mean & sd & median & min & max & skew & kurtosis \\ 
  \hline
original age & 748 & 9.16 & 6.89 & 10.50 & 0.04 & 21.18 & -0.03 & -1.56 \\ 
  synthetic age & 748 & 9.15 & 6.89 & 10.49 & 0.04 & 20.96 & -0.03 & -1.55 \\ 
  original hgt & 748 & 131.10 & 46.52 & 145.75 & 50.00 & 198.00 & -0.30 & -1.47 \\ 
  synthetic hgt & 748 & 131.06 & 46.50 & 145.32 & 50.69 & 197.16 & -0.30 & -1.47 \\ 
  original wgt & 748 & 37.12 & 26.03 & 34.55 & 3.14 & 117.40 & 0.38 & -1.03 \\ 
  synthetic wgt & 748 & 37.09 & 26.00 & 34.44 & 3.35 & 112.26 & 0.38 & -1.03 \\ 
  original bmi & 748 & 18.04 & 3.04 & 17.45 & 11.73 & 31.74 & 1.14 & 1.79 \\ 
  synthetic bmi & 748 & 18.05 & 3.08 & 17.48 & 11.49 & 32.37 & 1.11 & 1.85 \\ 
  original hc & 748 & 51.62 & 5.86 & 53.10 & 33.70 & 65.00 & -0.91 & 0.12 \\ 
  synthetic hc & 748 & 51.61 & 5.86 & 53.18 & 34.38 & 62.85 & -0.91 & 0.12 \\ 
  original gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.36 \\ 
  synthetic gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.35 \\ 
  original phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.25 \\ 
  synthetic phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.24 \\ 
  original tv & 748 & 8.43 & 8.12 & 3.00 & 1.00 & 25.00 & 0.85 & -0.78 \\ 
  synthetic tv & 748 & 8.42 & 8.11 & 3.19 & 1.00 & 25.00 & 0.85 & -0.77 \\ 
  original reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.77 \\ 
  synthetic reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.76 \\ 
   \hline
\end{tabular}
\end{table}

We see from Table 1 that the synthetic data estimates closely resemble the true data estimates. All sample statistics of interest show negligible bias over the 1000 synthetic data sets. Hence, univariately the synthetisation model proves adequate.

@@

An often used bivariate statistic is Pearson's correlation coefficient. When evaluating this correlation coefficient on the numeric columns in the `boys` data set, we find that biases are very small. The results are displayed in Table 3.

@@

However, when creating synthetic data with `mice`, close attention should be paid to three distinct aspects of the data: incorporating the additional uncertainty that is due to synthesizing (part of) the data, potential disclosure risks that remain after synthesizing the data, and the utility of the synthetic data. First, the procedure of generating multiple synthetic sets may seem overly complicated. We would like to emphasize that analyzing a single synthesized set, while perhaps unbiased, would underestimate the variance properties that are so important in drawing statistical inferences from finite data sets. After all, we are often not interested in the sample at hand, but aim to make inferences about the underlying data generating mechanism as reflected in the population. Properly capturing the uncertainty of synthetic data sets, just like with incomplete data sets, is therefore paramount. To achieve this, we adopted a bootstrap scheme in our simulations, to represent sampling from a superpopulation. However, when a single sample should be the reference, in the sense that the sample reflects the population one wishes to make inferences about, adjusted pooling rules are required that are similar to the procedure outlined in @vink2014pooling. The corresponding pooling rules have not been derived yet and the incorporation in data analysis workflows would require proper attention from developers alike.

@@

While `synthpop` is capable to synthesize regardless of missing data, it is not developed to solve missingness problems. If the data for synthetisation contains missing values, the missingness will be considered as a property of the data, and the corresponding synthetic data sets will contain missing values as well. If the goal is to solve for the missingness and synthesize the data set, a two-step approach must be used, in which the missing values are imputed using a different package, like `mice`. After this, `synthpop` can be used to synthesize the imputed data. Then, given $m$ multiple imputations and $r$ synthetisations, at least $m \times r$ synthetic data sets are in order, after which the data can be analysed and the analyses can be pooled using rules developed by @reiter2004simultaneous. However, such an approach would be computationally inefficient and practically cumbersome, because the architectures of both packages differ.

The appealing nature of `mice` therefore lays in its ability to solve both the missingness problem and create synthetic versions of the data. Eventually, the flexibility with `mice` is that both unobserved and observed data values could be synthesized at once, without the need for a two-step approach. Using `mice`, $m$ synthetic sets would be sufficient. However, as of today, no pooling rules for one-step imputation of missingness and synthetisation have been developed. The derivation of those would yield a massive theoretical improvement to further reduce the burden of creating synthetic data sets.

That said, the fields of synthetic data and multiple imputation are very much in development. Although efforts have been made to bridge gaps between these fields, such as in the `miceadds` package [@miceadds_3.11-6] which links functionality between between `synthpop` and `mice`, further improvements could be made in that respect. Specifically, `synthpop` contains a comprehensive suite of functions that can be used to assess the quality of synthetic data [e.g., @snoke2018general], while `mice` contains methodology to allow for flexible synthesis of intricate data structures, such as multilevel data. Being able to use such functionality interchangeably would greatly benefit data analysist.

>>>>

---
title: "Anony*mice*d shareable data: Using *mice* to create and analyze multiply imputed synthetic data sets"
author:
  - name: Thom Benjamin Volker
    affil: 1,*, \dagger
    orcid: 0000-0002-2408-7820
  - name: Gerko Vink
    affil: 1, \dagger
    orcid: 0000-0001-9767-1924
affiliation:
  - num: 1
    address: |
      Utrecht University - 
      Department of Methodology and Statistics. 
      Padualaan 14. 3584CH Utrecht, the Netherlands
# firstnote to eighthnote
firstnote: |
  These authors contributed equally to this work.
correspondence: |
  t.b.volker@uu.nl.
journal: psych
type: article
status: submit
bibliography: synth.bib
appendix: appendix.tex
abstract: |
  Synthetic data sets simultaneously allow for the dissemination of research data
  while protection the privacy and confidentiality of respondents. Generating 
  and analyzing synthetic data sets is straightforward, yet a synthetic data analysis 
  pipeline is seldom adopted by applied researchers. We outline a simple procedure 
  for generating and analyzing synthetic data sets with the multiple imputation
  software `mice` in `R`. We demonstrate using simulations that the analysis 
  results obtained on synthetic data yield unbiased and valid inferences, and 
  leads to synthetic records that cannot be distinguished from the true data records. 
  The ease of use when synthesizing data with `mice`, together with the validity of 
  inferences obtained through this procedure opens up a wealth of possibilities for 
  data dissemination and further research of initially private data. 
keywords: |
  mice; multiple imputation; synthetic data; statistical disclosure control; privacy.
acknowledgement: |
  We are grateful to Mirthe Hendriks and Stijn van den Broek for replicating the 
  simulation study on an independent data set. Additionally, we thank Hanne Oberman, 
  two anonymous reviewers and the editor for their helpful and insightful comments.
authorcontributions: |
  The authors contributed equally to this work.
conflictsofinterest: |
  The authors declare no conflict of interest. 
sampleavailability: |
  A full simulation archive is available from https://github.com/amices/Federated_imputation/tree/master/manuscript.
output: rticles::mdpi_article
---

@@

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values, one could easily replace any observed sensitive values by multiple draws from the posterior predictive distribution - or a non-Bayesian predictive distribution (both of these will be referred to as predictive distributions). Similarly to multiple imputation of missing data, the multiple synthetic data sets allow for correct statistical inferences, despite the fact that the analyses do not use the *true* value. The analyses over multiple synthetic data sets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure.

@@

Fully conditional specification (FCS) has been implemented in the `mice` package [@mice; @vanbuuren_fully_2006] in `R` [@Rproject], which has been developed for multiple imputation to overcome problems related to nonresponse. In that context, the aim is to replace missing values with plausible values from the predictive distribution of that variable. This is achieved by breaking down the multivariate distribution of the data $\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})$ into $j = 1, 2, \dots, k$ univariate conditional densities, where $k$ denotes the number of columns in the data. Using FCS, a model is constructed for every incomplete variable and the missing values $Y_{j, mis}$ are then imputed with draws from the predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ on a variable-by-variable basis. Note that the predictor matrix $Y_{-j}$ may contain yet imputed values from an earlier imputation step, and thus will be updated after every iteration. This procedure is applied $m$ times, resulting in $m$ completed data sets $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $\textbf{D}^{(l)} = (\textbf{Y}_{obs}, \textbf{Y}^{(l)}_{mis})$.

@@

The imputation approach in `mice` can be easily extended to generate synthetic values. Rather than imputing missing data, observed values are then replaced by synthetic draws from the predictive distribution. For simplicity, assume that the data is completely observed (i.e., $\textbf{Y} = \textbf{Y}_{obs}$). Following the notation of @reiter_raghunathan_multiple_2007, let, given $n$ observations, $Z_i = 1$ if any of the values of unit $i = 1, 2, \dots, n$, are to be replaced by imputations, and $Z_i = 0$ otherwise, with $Z = (Z_1, Z_2, \dots, Z_n)$. Accordingly, the data consists of values that are to be replaced and values that are to be kept (i.e., $\textbf{Y} = (\textbf{Y}_{rep}, \textbf{Y}_{nrep})$. Now, instead of imputing $\textbf{Y}_{mis}$ with draws from the predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ as in the missing data case, $\textbf{Y}_{rep}$ is imputed from the posterior distribution of $P(Y^{(l)}_{j, rep} | \textbf{Y}_{-j}, Z, \theta)$, where $l$ is an indicator for the synthetic data set ($l = 1, 2, \dots, m$). This process results in the synthetic data $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$.

@@

```{r eval = FALSE}
syn <- mice(data.to.synthesize, 
            m = 10,
            maxit = 1,
            where = matrix(TRUE,
                           nrow = nrow(data.to.synthesize),
                           ncol = ncol(data.to.synthetize)))
```

Here, the argument `where` requires a matrix of the same dimensions as the data, (i.e., a $n \times k$ matrix) containing logicals $z_{ij}$ that indicate which cells are selected to have their values replaced by draws from the predictive distribution. This approach allows to replace a subset of the observed data (e.g., by specifying only those cells that are to be replaced as `TRUE` in the `where`-matrix, leaving the rest to `FALSE`), or as in the aforementioned example, the observed data as a whole, resulting in a data set that partially or completely consists of synthetic data values. Note that because the data is completely observed, iterating over the predictive distribution is not required.

@@

Choosing an adequate imputation model to impute the data is paramount, as a flawed imputation model may drastically impact the validity of inferences [@grund2021using; @jiang2021balancing]. Imputation models should be as flexible as possible to capture most of the patterns in the data, and to model possibly unanticipated data characteristics [@murray_multiple_2018; @rubin_18years_1996]. Generally, multiple methods can be feasible, including joint modeling, sequential regression and FCS, all with different parameter specifications, depending on the imputation problem at hand [@ludtke2020regression; @yucel2018sequential; @murray2016multiple]. Various simulations have shown that the most accurate imputation procedure may vary over different imputation problems [e.g., see @hanne_review_2021 for an overview]. Hence, a one-size-fits-all solution does probably not exist.

However, one of the methods that has shown a reasonable and consistent performance over multiple simulations [e.g., @reiter_cart_2005; @burgette_reiter_cart_2010; @doove_buuren_recursive_2014; @raab_practical_2016] is FCS in combination with classification and regression trees [CART\; @breiman_cart_1984]. Due to its non-parametric nature, CART allows to model complex patterns (e.g., interactions and non-linear relationships) in the data. Loosely speaking, CART sequentially splits the predictor space into non-overlapping regions in such a way that the within-region variance is as small as possible after every split. As such, CART does not impose any parametric distribution on the data, making it a widely applicable method that allows for a large variety of relationships within the data to be modeled [@islr_2013]. Given these appealing characteristics and the call for the use of flexible methods when multiply imputing data, we will focus our illustrations and evaluations of `mice` to method `mice.impute.cart()`, realized by:

@@

In a nutshell, the above code shows the simplicity of creating $m = 10$ synthetic data sets using `mice`. In practice, however, one should take some additional complicating factors into account. For example, one should account for deterministic relations in the data. Additionally, relations between variables may be described best using a different model than `CART`. Such factors are data dependent and should be considered by the imputer. In the next section, we will describe how a completely observed version of the `mice::boys` data [@fredriks_boys_2000] can be adequately synthesized. Additionally, we will show through simulations that this approach yields valid inferences.

@@

Unfortunately, this data set does not differ from the vast majority of collected data sets, in the sense that it suffers from missing data. For simplicity, we solve the missingness problem using a single imputation with the default `mice` imputation model for all predictors except `bmi`, which is passively imputed using its deterministic relation with `wgt` and `hgt`. Specifically, the imputed values are used to calculate the exact `bmi` values that correspond with `hgt` and `wgt`.

@@

Usually, one would draw samples from a population that can be synthesized to evaluate the performance of the synthetisation methods. As we only have access to a single sample, 1000 bootstrap samples of the `boys` data have been synthesized with $m = 5$ imputations for every data cell to induce an appropriate amount of sampling variance. This approach precludes us from knowing the *true* data-generating model, and in this sense provides a more stringent test of the applicability of `mice` for real-life synthetisation procedures. Specifically, every bootstrapped sample is treated as an actual sample from a population. These bootstrap samples will be synthesized using `mice`, using a model that is built to approximate the true data-generating mechanisms as closely as possible. As relationships in realistic data are generally more complex than relationships in parametrically simulated data, achieving good performance on a real data set is likely to be more indicative of practical applicability than good performance on samples from a known multivariate probability density function.

@@

$$
\bar{u}_m = \sum_{l = 1}^m \frac{u^{(l)}}{m},
$$
where $u^{(l)}$ denotes the variance estimate in the $l$th synthetic data set. These pooling rules are implemented in `mice` as the function `pool.syn`. @reiter_kinney_2012 proved that synthesizing data from the posterior predictive distribution is not required for these pooling rules to hold. In fact, the parameters of an imputation model can be fixed to their maximum likelihood estimates or their posterior modes, and synthetic values can be drawn after plugging these values into the imputation models. This observation as been taken a step further by @raab_practical_2016, who developed pooling rules that do not require multiple synthetic data sets and performed highly similar to the aforementioned pooling rules. However, because these pooling rules can only be used when the complete data (rather than a subset of the data) is synthesized, we adhere to the originally proposed pooling rules by @reiter_partially_inference_2003, who defined the total variance in terms of sampling variance and between-imputation variance.

@@

The synthetic data is evaluated with respect to the *true* data set on the basis of three aims. We believe that every reliable and valid data synthetisation effort in statistical data analysis should be able to yield 1) unbiased univariate statistical properties, 2) unbiased bivariate properties, 3) unbiased and valid multivariate inferences and 4) synthetic data that cannot be distinguished from real data. We consider the evaluation of the synthetic data simulations in the above order.

@@

The univariate descriptives for the original data and the synthetic data can be found in Table 2. From this table, it can be seen that the synthetic data estimates closely resemble the true data estimates. All sample statistics of interest show negligible bias over the 1000 synthetic data sets.

@@

Both figures show that the marginal distributions of the bootstrapped data and the synthetic data are highly similar. The sampling distributions of these parameters hardly seem to differ. Hence, both the numeric and graphical results show that univariately the synthetisation model proves adequate.

@@

```{r univariate-statistics-II, echo = FALSE, cache = TRUE, results = FALSE, fig.cap = "Empirical sampling distribution of the number of observations within each category of the categorical variables in the bootstrapped and synthetic boys data."}
boot_stats_cat <- bootstrap_samples %>%
  map_dfr(~.x %>%
            select(!where(is.numeric)) %>%
            mutate(across(everything(), as.character)) %>%
            pivot_longer(everything()) %>% 
            group_by(name, value) %>%
            summarize(`Number of observations` = n(), .groups = "drop"))

@@

```{r, include = FALSE}
library(xtable)
bind_rows(true, setNames(synth, names(true))) %>% 
           arrange(vars) %>% .[, c(2:5, 8:9, 11:12)] %>% round(3) %>% 
  xtable
```
<!-- % latex table generated in R 4.1.0 by xtable 1.8-4 package -->
<!-- % Mon Oct  4 16:47:59 2021 -->
\begin{table}[ht]
\caption{Univariate descriptives for the true data and $m=5$ pooled univariate descriptives for the synthetic data over 1000 simulations. Variable names followed by an asterix ($^*$) are categorical.}
\centering
\begin{tabular}{rrrrrrrrr}
  \hline
 & n & mean & sd & median & min & max & skew & kurtosis \\ 
  \hline
original age & 748 & 9.16 & 6.89 & 10.50 & 0.04 & 21.18 & -0.03 & -1.56 \\ 
  synthetic age & 748 & 9.15 & 6.89 & 10.49 & 0.04 & 20.96 & -0.03 & -1.55 \\ 
  original hgt & 748 & 131.10 & 46.52 & 145.75 & 50.00 & 198.00 & -0.30 & -1.47 \\ 
  synthetic hgt & 748 & 131.06 & 46.50 & 145.32 & 50.69 & 197.16 & -0.30 & -1.47 \\ 
  original wgt & 748 & 37.12 & 26.03 & 34.55 & 3.14 & 117.40 & 0.38 & -1.03 \\ 
  synthetic wgt & 748 & 37.09 & 26.00 & 34.44 & 3.35 & 112.26 & 0.38 & -1.03 \\ 
  original bmi & 748 & 18.04 & 3.04 & 17.45 & 11.73 & 31.74 & 1.14 & 1.79 \\ 
  synthetic bmi & 748 & 18.05 & 3.08 & 17.48 & 11.49 & 32.37 & 1.11 & 1.85 \\ 
  original hc & 748 & 51.62 & 5.86 & 53.10 & 33.70 & 65.00 & -0.91 & 0.12 \\ 
  synthetic hc & 748 & 51.61 & 5.86 & 53.18 & 34.38 & 62.85 & -0.91 & 0.12 \\ 
  original gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.36 \\ 
  synthetic gen* & 748 & 2.53 & 1.59 & 2.00 & 1.00 & 5.00 & 0.52 & -1.35 \\ 
  original phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.25 \\ 
  synthetic phb* & 748 & 2.75 & 1.86 & 2.00 & 1.00 & 6.00 & 0.56 & -1.24 \\ 
  original tv & 748 & 8.43 & 8.12 & 3.00 & 1.00 & 25.00 & 0.85 & -0.78 \\ 
  synthetic tv & 748 & 8.42 & 8.11 & 3.19 & 1.00 & 25.00 & 0.85 & -0.77 \\ 
  original reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.77 \\ 
  synthetic reg* & 748 & 3.02 & 1.14 & 3.00 & 1.00 & 5.00 & -0.08 & -0.76 \\ 
   \hline
\end{tabular}
\end{table}

@@

An often used bivariate statistic is Pearson's correlation coefficient. When evaluating this correlation coefficient on the numeric columns in the `boys` data set, the differences between the correlations in the synthetic and bootstrapped data are very small. These results are displayed in Table 3.

@@

When creating synthetic data with `mice`, close attention should be paid to three distinct factors: (i) the additional uncertainty that is due to synthesizing (part of) the data should be incorporated, (ii) potential disclosure risks that remain after synthesizing the data should be assessed, and (iii) the utility of the synthetic data should be examined. First, the procedure of generating multiple synthetic sets may seem overly complicated. We would like to emphasize that analyzing a single synthesized set, while perhaps unbiased, would underestimate the variance properties that are so important in drawing statistical inferences from finite data sets. After all, we are often not interested in the sample at hand, but aim to make inferences about the underlying data generating mechanism as reflected in the population. Properly capturing the uncertainty of synthetic data sets, just like with incomplete data sets, is therefore paramount. To achieve this, we adopted a bootstrap scheme in our simulations, to represent sampling from a superpopulation. However, when a single sample should be the reference, in the sense that the sample reflects the population one wishes to make inferences about, adjusted pooling rules are required that are similar to the procedure outlined in @vink2014pooling. The corresponding pooling rules have not been derived yet and the incorporation in data analysis workflows would require proper attention from developers alike.

@@

While `synthpop` is capable to synthesize regardless of missing data, it is not developed to solve missingness problems. If the data for synthetisation contains missing values, the missingness will be considered as a property of the data, and the corresponding synthetic data sets will contain missing values as well. If the goal is to solve for the missingness and synthesize the data set, a two-step approach must be used, in which the missing values are imputed using a different package, like `mice`. Subsequently, `synthpop` can be used to synthesize the imputed data. Then, given $m$ multiple imputations and $r$ synthetisations, at least $m \times r$ synthetic data sets are in order, after which the data can be analysed and the analyses can be pooled using rules developed by @reiter2004simultaneous. However, such an approach would be computationally inefficient and practically cumbersome, because the architectures of both packages differ.

The appealing nature of `mice` therefore lays in its ability to both solve the missingness problem and create synthetic versions of the data. Eventually, the flexibility with `mice` is that unobserved and observed data values could be synthesized at once, without the need for a two-step approach. Using `mice`, $m$ synthetic sets would be sufficient. However, as of today, no pooling rules for one-step imputation of missingness and synthetisation have been developed. The derivation of those would yield a massive theoretical improvement to further reduce the burden of creating synthetic data sets.

That said, the fields of synthetic data and multiple imputation are very much in development. Although efforts have been made to bridge gaps between these fields, such as in the `miceadds` package [@miceadds_3.11-6] which links functionality between between `synthpop` and `mice`, further improvements could be made in that respect. Specifically, `synthpop` contains a comprehensive suite of functions that can be used to assess the quality of synthetic data [e.g., see @snoke2018general], while `mice` contains methodology to allow for flexible synthesis of intricate data structures, such as multilevel data. Being able to use such functionality interchangeably would greatly benefit data analysts.
